\section{Case study: “Do a peatland site survey on the forest of Bowland” (MCP-Geo)}

\subsection{Scope and evidence (primary)}

This case study is reconstructed from four artefacts, treated as \textbf{primary evidence} \cite{artefact_claude_stopped_after_ngd_features_2026_02_18,artefact_codex_trace_jsonl_2026_02_18,artefact_floor_question_analysis_2026_02_19,artefact_forensic_study_mcp_geo_2026_02_19}:

\begin{itemize}[leftmargin=2em]
  \item Claude transcript (floor question, tool excerpts, adapter truncation, final stop marker). \cite{artefact_claude_stopped_after_ngd_features_2026_02_18}
  \item Tool trace (\texttt{codex-trace.jsonl}; JSONL of MCP tool calls). \cite{artefact_codex_trace_jsonl_2026_02_18}
  \item Trace-derived deep analysis and timeline. \cite{artefact_floor_question_analysis_2026_02_19}
  \item Forensic narrative and engineering recommendations. \cite{artefact_forensic_study_mcp_geo_2026_02_19}
\end{itemize}

The initiating floor question is:
\begin{quote}
“Do a peatland site survey on the forest of Bowland”. \cite{artefact_claude_stopped_after_ngd_features_2026_02_18}
\end{quote}

A second constraint introduced mid-session was:
\begin{quote}
“Don't use the web, use mcp-geo”. \cite{artefact_claude_stopped_after_ngd_features_2026_02_18}
\end{quote}

\subsection{Narrative reconstruction}

\subsubsection*{Discovery attempt (correct tool, but perceived failure)}
\begin{itemize}[leftmargin=2em]
  \item The workflow started with an \textbf{OS Names} lookup for “Forest of Bowland”
    (\texttt{os\_names\_find}, \texttt{id=15}) and received a \textbf{successful} response.
  \item The transcript records a \textbf{user-visible} client/UI error (“No result received from client-side tool execution.”),
    creating an early mismatch between what the MCP server did and what the host presented.
\end{itemize}

\noindent$\rightarrow$ \hangindent=1.8em \hangafter=1 \textbf{Consequence:} the workflow pivoted away from the “right” discovery path into address/POI approaches.

\subsubsection*{Boundary/extent resolution failed (protected landscape not available)}
\begin{itemize}[leftmargin=2em]
  \item An administrative lookup for “Forest of Bowland” returned \textbf{no match}
    (admin lookup, \texttt{id=16}).
  \item A later admin lookup for “Bowland” returned an \textbf{electoral ward} (\texttt{id=19}).
    The ward polygon was then pulled (\texttt{admin\_lookup\_area\_geometry}, \texttt{id=20}).
\end{itemize}

\noindent$\rightarrow$ \hangindent=1.8em \hangafter=1 \textbf{Consequence:} the “area of interest” became \textbf{the ward boundary}, not the \textbf{Forest of Bowland AONB / National Landscape} boundary.

\subsubsection*{NGD feature interrogation escalated (schema → filters → geometry)}
\begin{itemize}[leftmargin=2em]
  \item Collection discovery and queryables/schema discovery were used effectively (\texttt{id=21--22}).
  \item The workflow attempted “peat” detection by filtering on land-cover descriptions (\texttt{id=23}, \texttt{id=24}, then broader CQL in \texttt{id=25}).
  \item The CQL request timed out upstream (\texttt{id=25}, status 501; \texttt{read timeout=5}).
    The workflow then continued with narrower filters (\texttt{id=26}, \texttt{id=27}).
\end{itemize}

\noindent$\rightarrow$ \hangindent=1.8em \hangafter=1 \textbf{Consequence:} the workflow shifted into larger “sample pulls” and then geometry-heavy calls (\texttt{id=28} then \texttt{id=30}), increasing payload size and host pressure.

\subsubsection*{Degradation → stop (payload + latency + host truncation)}
\begin{itemize}[leftmargin=2em]
  \item A large geometry-bearing land response occurred (\texttt{id=30}, response \textasciitilde{}555 kB).
  \item A final geometry-bearing water query ran for a long time (\texttt{id=31}, latency \textasciitilde{}317 s; response \textasciitilde{}271 kB).
\end{itemize}

\noindent$\rightarrow$ \hangindent=1.8em \hangafter=1 \textbf{Consequence:} the experience looked like “the agent stopped”, even though the trace shows the final tool call returned HTTP 200.

\subsection{Timeline table (tool discovery → calls → degradation → stop)}

All timestamps are \textbf{UTC}. \cite{artefact_codex_trace_jsonl_2026_02_18,artefact_floor_question_analysis_2026_02_19,artefact_forensic_study_mcp_geo_2026_02_19}

\begin{itemize}[leftmargin=*]
  \item \textbf{15 (14:34:37)} OS Names lookup succeeded, but the host reported “no result”.
  \item \textbf{16 (14:35:11)} Admin lookup found no match for “Forest of Bowland”.
  \item \textbf{19--20 (14:45:26--14:45:35)} “Bowland” resolved to a ward; ward geometry used as proxy AOI.
  \item \textbf{21--22 (14:45:47--14:46:40)} Land collections and schema/queryables discovery succeeded.
  \item \textbf{25 (14:47:40)} Upstream timeout (status 501; read timeout=5).
  \item \textbf{30 (14:49:40)} Large geometry payload returned (\textasciitilde{}555 kB).
  \item \textbf{31 (14:50:24)} Very long call (\textasciitilde{}317 s) followed by truncation/stop.
\end{itemize}

\begin{figure}[ht]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
  x=1cm,y=1cm,
  font=\scriptsize,
  >=Latex,
  event/.style={draw,rounded corners,fill=gray!6,inner sep=3pt,align=left},
  call/.style={draw,rounded corners,fill=blue!6,inner sep=3pt,align=left},
  fail/.style={draw,rounded corners,fill=red!6,inner sep=3pt,align=left},
  heavy/.style={draw,rounded corners,fill=orange!10,inner sep=3pt,align=left},
  axis/.style={draw,thick,-{Latex[length=3mm]}}
]
% timeline axis (taller canvas and explicit break)
\draw[axis] (0,0) -- (16.8,0) node[right]{time};

% ticks (compressed after last call)
\foreach \x/\t in {0/14:34, 2/14:35, 7/14:45, 10/14:46, 12/14:47, 14/14:49, 15.2/14:50} {
  \draw (\x,0.12) -- (\x,-0.12);
\node[below] at (\x,-0.12) {\scriptsize \t};
}
% explicit time-break marker (not to scale)
\draw[densely dashed,thick] (15.95,-0.45) -- (15.95,2.25);
\node[fill=white,inner sep=1pt] at (15.95,0.85) {\scriptsize time break};
\node at (15.8,0) {\normalsize$\cdots$};

% events
\node[call] (e15) at (0,1.65) {\textbf{id=15}\\OS Names\\\textit{host: ``no result''}};
\draw[-{Latex[length=2.5mm]}] (0,0.15) -- (e15.south);

\node[event] (e16) at (2,1.65) {\textbf{id=16}\\Admin lookup\\\textit{no match}};
\draw[-{Latex[length=2.5mm]}] (2,0.15) -- (e16.south);

\node[event] (e19) at (7,1.65) {\textbf{id=19--20}\\Ward proxy AOI\\resp \textasciitilde{}46.9 kB};
\draw[-{Latex[length=2.5mm]}] (7,0.15) -- (e19.south);

\node[call] (e21) at (10,1.65) {\textbf{id=21--22}\\Discovery+schema\\resp \textasciitilde{}26.7/41.5 kB};
\draw[-{Latex[length=2.5mm]}] (10,0.15) -- (e21.south);

\node[fail] (e25) at (12,1.65) {\textbf{id=25}\\Timeout\\501; 15.6 s};
\draw[-{Latex[length=2.5mm]}] (12,0.15) -- (e25.south);

\node[heavy] (e30) at (14,1.65) {\textbf{id=30}\\Large geometry\\resp \textasciitilde{}555 kB};
\draw[-{Latex[length=2.5mm]}] (14,0.15) -- (e30.south);

\node[heavy] (e31) at (15.2,1.65) {\textbf{id=31}\\Long call\\\textasciitilde{}317 s; resp \textasciitilde{}271 kB\\\textit{truncation/stop}};
\draw[-{Latex[length=2.5mm]}] (15.2,0.15) -- (e31.south);

% connectors
\draw[->,thin] (e15.east) -- (e16.west);
\draw[->,thin] (e16.east) -- (e19.west);
\draw[->,thin] (e19.east) -- (e21.west);
\draw[->,thin] (e21.east) -- (e25.west);
\draw[->,thin] (e25.east) -- (e30.west);
\draw[->,thin] (e30.east) -- (e31.west);
\end{tikzpicture}%
}
\caption{Compressed timeline with payload/latency cues (not to scale after 14:50; ellipsis indicates compression).}
\end{figure}

\subsection{Confirmed turning points (discovery → failure → degradation → stop)}

Evidence for turning points is taken from the transcript and tool trace. \cite{artefact_claude_stopped_after_ngd_features_2026_02_18,artefact_codex_trace_jsonl_2026_02_18}

\begin{enumerate}[leftmargin=*]
  \item \textbf{Discovery (but user-visible failure):} \texttt{id=15} succeeded, yet the host indicated “no result received”.
  \item \textbf{Boundary failure:} \texttt{id=16} returned no match; \texttt{id=19–20} fell back to a ward polygon.
  \item \textbf{Degradation:} schema fetch (\texttt{id=22}), upstream timeout (\texttt{id=25}), then geometry-heavy calls (\texttt{id=30}, \texttt{id=31}).
  \item \textbf{Stop (user-visible):} adapter truncation appears before the final “could not be fully generated” marker.
\end{enumerate}
